A state space model reveals the rich and varied dynamics of replay

### 2020-07-01 Article.tex
Update on Overleaf.

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\DeclareMathOperator*{\E}{\mathbb{E}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

@@

\title{A state space model for characterizing the trajectory dynamics of replay}
\shorttitle{Replay Dynamics}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past or possible future experiences play a critical role in decision-making processes. The hippocampus is a brain region known to express these representations across multiple states, including during sharp-wave ripple (SWR) events where sets of hippocampal place cells can be activated in time-compressed sequences consistent with a rapid mental traversal of a path. Efforts to understand these representations typically make assumptions about the character and dynamics of the generative representations. As a result, we lack a clear picture of the nature and diversity of SWR representations. Here we develop a flexible state space model that uses a combination of discrete and latent states to describe the content and time evolution of SWR events. Application of this model reveals that a much larger proportion of SWR events contain coherent spatial content that previously reported, and that the dynamics of events span a surprisingly wide range.

@@

\section*{Introduction}
The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These internal "generative" representations can be identified because the underlying neural activity patterns resemble those seen during actual events. In particular, an encoding model that relates neural firing to external covariates can be constructed during an experience, and this model can inverted to decode an internal representation of the covariates from the neural activity at a different time.

Mental events do not occur in the same time span as actual events; memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly. As a result, decoding generative representations requires making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the stationary state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
In order to introduce and validate the model, we demonstrate how the model works on a simulated dataset (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly, representing a single location. For the middle third of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the dynamics model--how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we use small time bins (2 ms vs. 20 ms) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a dynamics model for how latent position---the "mental" position of the animal represented by the cells---can evolve in each of these sequences (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can change from the previous time step. In the stationary dynamic, we expect latent position to not change between time steps, so we use an identity state transition matrix, which predicts the next position will be the same as the last position. In the continuous dynamic, we expect the latent position to be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since we are interested in identifying replays that move at speeds much faster than the animal's speed, we set this to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or 24.5 m/s), consistent with replay speeds observed in previous studies \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}. In the fragmented dynamic, we expect that the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes all positions equally likely.

Finally, we specify how likely the dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for the duration of the SWR, with a small probability of switching to one of the other dynamics. Therefore, we set the probability of staying in a state to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular state.

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic state. With this, we can summarize the resulting posterior with two quantities: the mixture probability of each dynamic over time and the probability of position over time, irrespective of dynamic (Figure \ref{1}F, left and right plot respectively). From these two summaries, we can see that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The probability of the stationary dynamic has high probability at the beginning of the simulated replay, reflecting the consistent spiking from one cell. This gives way to the continuous dynamic, reflecting the trajectory-like spiking from the population of cells. Then the fragmented dynamic dominates for the last third when the population spiking is not spatially coherent. We can also see that these three dynamics are reflected in the estimate of latent position of the animal.  Importantly, we get a much less variable estimate due to the correct specification of the dynamics.

One way to interpret the probabilities in our model is as a classifier of speeds of replay. To demonstrate this, we applied the model to simulated trajectories of increasing speed (Figure \ref{1}G), from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three dynamics being high probable (where we define highly probable to be over 0.8 probability), there are in-between speeds where two of the dynamics dominate with high probability; that is, the sum of two of the dynamics' probabilities is over 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary state has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its dynamics.

In order to verify that the model was robust to the choice of probability of staying in a state, we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure 1-Supplemental Figure 1). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v4}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes (arranged top to bottom). In the top panel, 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center wells respectively. The animal's position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} 1D MAP estimate of the latent position in (A) projected back into 2D. Color indicates time. The animal's position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center well as in (A). \textbf{(D)} 1D MAP estimate of the latent position in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{The model finds known continuous replays with sorted and clusterless spikes}

@@

Next, we wanted to validate the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with clear sequential population activity (Figure \ref{2}A, top panel). We observed that the probability of being in the continuous dynamic is high throughout for this SWR, with the probability of being in a stationary state being slightly higher at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory to 2D (Figure \ref{2}C) to better see the spatial trajectory. Moreover, when we apply the same model using the 2D position of the animal, we get a similar result as when we used the 1D linearized position (Figure 2-Supplemental 1A).

One problem with using spikes from cells identified using spike sorting is that there is potentially a lot of information lost by determining the identity of each cell---many spikes are discarded because the spikes could not be uniquely assigned to one cell. Therefore, we wanted to use a method which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding. Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}). Here, if we apply it to the same SWR as the sorted spikes, we get similar decoding and classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure 2-Supplemental 1B). Note that, using clusterless decoding, the estimate of the posterior probability of position is much more narrow, indicating that there is more certainty in the true latent position at that time compared the corresponding posterior probability of position using sorted spikes (Figure \ref{2}D vs C).

\subsection*{The model finds replays with non-constant speed dynamics}
After validating our model on a single SWR, we then applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009}). We observed many replays that were classified as continuous throughout the entirety of SWR and likely would have been identified by the typical Bayesian Decoder (Figure 2-Supplemental 2A-C). However, we also observed replays that did not have this structure, but could be interpreted in terms of our model. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure 3-Supplemental 1B), trajectories that stayed in one position (Figure \ref{3}B), trajectories that jumped between arms (Figure \ref{3}C-D, Figure3-Supplemental 1A, E, F), and trajectories that remained spatially incoherent throughout the SWR (Figure3-Supplemental1D). Overall, we were able to classify 89\% ( 23382 of 26160) of SWRs as containing at least one of the five dynamics. To check that this depended on the spatial receptive fields of cells in the hippocampus, we resampled position with replacement for two epochs, breaking the relationship between spiking and spatial information. There were far fewer SWR classified when spatial information was removed (99\% true classified, 9\% median resampled distribution, p=0.02 for epoch 1, p=0.02 for epoch 2, Figure 3-Supplemental 2).

To better understand the distribution of dynamics during SWRs, we then asked what percentage of classified replays contained spatially coherent content---which we define as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics. We found that 22478 of 23382 or 96\% of SWR had some spatially coherent structure while 3353 of 23382 or 14\% had some spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. We found that, consistent with previous reports, only 4511 of  23382 or 19\% of SWRs had some continuous content (Figure \ref{3}G). This highlights that much of the SWR replay content is missed when analyzing SWRs with the Bayesian decoder.

To verify that partitioning the data into spatially coherent and incoherent time segments by dynamic was meaningful, we quantified the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. For example, Figure \ref{4}A and B show two SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated on a small portion of the track and so the cumulative HPD is low throughout the SWRs. In contrast, the Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. This can change over the time course of a SWR, as in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher. When we looked at the average cumulative HPD for each SWR by dynamic, we found a clear bi-modal split between spatially congruent dynamics and spatially incongruent dynamics (Figure \ref{4}E). For the spatially congruent dynamics, the cumulative HPD was much lower than the spatially incongruent dynamics (median 24 cm, spatially congruent vs. median 238 cm, spatially incongruent).

@@

We next looked at which sets of dynamics are most common. Most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were ones that were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These trajectories had a slightly shorter duration to events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B) and occurred at a similar average distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's position and move at a speed that is closer to the animal's actual movement speed when running, rather than many times the animal's movement speed.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these trajectories represented the animal's position (Figure \ref{5}C), but there were many that represented positions as far as 226 cm from the animal. We define the ones that are more than 30 cm away from the animal as "non-local stationary trajectories". There were 1585 of 23382 classified SWRs or 7\% non-local stationary trajectories and 731 of these non-local stationary trajectories were stationary throughout the duration of the SWR. These non-local stationary trajectories represented positions all over the track, but were most common at reward wells and choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}. The average multiunit firing rate during stationary dynamics was not lower than the other dynamics (Figure \ref{5}E.

@@

>>>>

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\DeclareMathOperator*{\E}{\mathbb{E}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

@@

\title{A state space model reveals the rich and varied dynamics of replay}
\shorttitle{Replay Dynamics}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past or possible future experiences play a critical role in memory and decision-making processes. The hippocampus is a brain region known to express these representations across multiple states, including during sharp-wave ripple (SWR) events where sets of hippocampal place cells can be activated in time-compressed sequences consistent with a rapid mental traversal of a path. Efforts to understand these "replay" events typically analyze a small subset of events and assume rapid mental movement, and as a result we lack a clear picture of both the prevalence and the range of dynamics associated with replay. Here we develop a flexible state space model that uses a combination of discrete and latent states to capture the content and time evolution of SWR events. Application of this model reveals the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

\section*{Introduction}
The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These internal "generative" representations \cite{KayConstantSubsecondCycling2020} can be identified because the underlying neural activity patterns resemble those seen during actual events. In particular, an encoding model that relates neural firing to external covariates can be constructed during an experience, and this model can inverted to decode an internal representation of the covariates from the neural activity at a different time.

Mental events may not occur in the same time span as actual events; memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly. As a result, decoding generative representations often involves  making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

@@

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the stationary state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition. (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly, representing a single location. For the middle third of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the dynamics model--how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a dynamics model for how latent position---the "mental" position of the animal represented by the cells--- evolves over time (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

In the stationary dynamic, the latent position would not change between time steps, so we use an identity state transition matrix, which predicts the next position will be the same as the last position. In the continuous dynamic, the latent position would be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since we are interested in identifying replays that move at speeds much faster than the animal's speed, we set this to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (corresponding to a latent speed of up to ~25 m/s). This is consistent with replay speeds observed in previous studies \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}. Finally, in the fragmented dynamic, the latent position could move to any available position instantaneously. We model this using a uniform state transition matrix, which makes all positions equally likely.

Finally, we specify how likely the dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for the duration of the SWR, with a small probability of switching to one of the other dynamics (note that  below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a state to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular state.

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic state (see methods). With this, we can summarize the resulting posterior with two quantities: the mixture probability of each dynamic over time and the probability of position over time, irrespective of dynamic (Figure \ref{1}F, left and right plot respectively).

An examination of these two summaries reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and the probability of the stationary dynamic is high at the beginning of the simulated replay. The change in the character of the data then drives a rapid transition to the continuous dynamic, reflecting the trajectory-like spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, and as a result, the fragmented dynamic dominates for the last third. Importantly, these dynamics also inform the estimate of latent position of the animal, and the correct assignment of dynamic reduces the variance of the latent position estimate.

The specification of these dynamics allows the model to capture a wide range of movement speeds for the latent position.  To demonstrate this, we applied the model to simulated trajectories of increasing speed (Figure \ref{1}G), from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three dynamics being high probable (where we define highly probable to be over 0.8 probability), there are intermediate speeds where two of the dynamics dominate with high probability; that is, the sum of two of the dynamics' probabilities is over 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary state has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its dynamics.

We note here that the choice of any particular threshold is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95 because the 0.95 threshold often leads people to conclude that only events that are statistically significant at the 0.05 level are likely to be meaningful to the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

As mentioned above, the model is robust to the choice of probability of staying in a state. We decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure 1-Supplemental Figure 1). We found that for a large range of plausible probabilities of staying in one of the dynamics (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v4}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} 1D MAP estimate of the latent position in (A) projected back into 2D. Color indicates time. The animal's actual position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center reward well as in (A). \textbf{(D)} 1D MAP estimate of the latent position in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

@@

Our success with simulated data inspired a further validation of the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this SWR, but the probability of being in a stationary state was noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory to 2D (Figure \ref{2}C) to better see the spatial trajectory. Moreover, when we apply the same model using the 2D position of the animal, we get a similar result as when we used the 1D linearized position (Figure 2-Supplemental 1A).

One of our criteria for a more optimal method is that it use all of the available spiking data. Using only clustered spikes discards all of the spike events that cannot be uniquely assigned to a putative single neuron, reducing very substantially the quality of the resulting decoding. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding (see Methods). Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}). Applying a clusterless decoder to the same SWR, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure 2-Supplemental 1B). The spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model, however. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event, and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

\subsection*{Replays with non-constant speed}

After testing our model on a single SWR, we applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020}). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a relatively permissive threshold (see methods) to ensure that we were included both the rare large events as well as the much more common small events.

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of SWR (Figure 2-Supplemental 2A-C). However, we also observed many events with spatially coherent content that did not have this structure. For example, there were trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure 3-Supplemental 1B), trajectories that remained fixed in one position (Figure \ref{3}B), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-D, Figure3-Supplemental 1A, E, F), and trajectories that were spatially incoherent throughout the SWR (Figure3-Supplemental1D).

 Using our 0.8 threshold, we were able to classify 89\% ( 23382 of 26160) of SWRs as containing at least one of the five dynamics. To ensure that this depended on the spatial receptive fields of cells in the hippocampus, we resampled position with replacement for two epochs, shuffling the relationship between spiking and spatial information. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for epoch 1, p=0.02 for epoch 2, Figure 3-Supplemental 2).

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content [REFS]. We therefore asked what fraction of events contained spatially coherent content in any form, defined as containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see methods). We found that 22478 of 23382 or 96\% of SWRs had some spatially coherent structure while 3353 of 23382 or 14\% had some spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of  23382 or 19\% of SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent but not continuous content.

Our model is specified in the context of a marked point process associated with different dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

We found that spatially coherent events did indeed have lower cumulative HPDs. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and  the cumulative HPD is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD for each SWR, grouped by dynamic, we found a clear bi-modal split between spatially congruent dynamics and spatially incongruent dynamics (Figure \ref{4}E). For the spatially congruent dynamics, the cumulative HPD was much lower than the spatially incongruent dynamics (median 24 cm, spatially congruent vs. median 238 cm, spatially incongruent).

@@

Our results indicate that most events with coherent spatial content are not continuous. We therefore examined these events in more details. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, 
than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events  represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these events represented a location close to the the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events  that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify "non-local stationary trajectories" and found these trajectories in 1585 of 23382 (~7\%) of classified SWRs. About half of these (731)  were stationary throughout the duration of the SWR. These non-local stationary trajectories most commonly represented  reward wells or choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}. 
Finally, we examined the overall activity levels across dynamics and found similar total spiking rates, indicating ...

 (Figure \ref{5}E.

@@

relationship to Johnson and Redish 2008

extension to previously experienced tracks
